\paperTitle

### 2014-05-10 doc/icml2014/background.tex
Various minor corrections to camera-ready

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
Let $\Pa(a) \subset X \union H$ be the parents of a variable $a$ in a directed
  model, and $\sN(a) \subset X \union H$ be the neighbors of $a$ in an
  undirected model.
We will use $x_1, x_2, \cdots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For an integer $n$, let $[n] = \{ 1, \dots, n \}$.
For ease of exposition, we assume that $x_v \in [d]$ for
all $v$, and that $h_i \in [k]$ for all $i$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef [d]^L$ and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef [k]^M$.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\vec
  i = (i_1, \cdots, i_\ell)$, define the projection $T[\vec i] = T_{i_1 \ldots i_\ell}$.
Similarly, for a set $\sC$ of size $k$ and vector $\vec x \in \Re^d$, we
  will use the notation $\vec x_\sC$ to get a vector in $\Re^k$ which selects
  the appropriate elements of $\vec x$.

We use $\otimes$ to denote the tensor product: if $u \in \Re^d,
  v \in \Re^k$, then $u \otimes v \in \Re^{d \times k}$.
For an $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $x_1, \cdots, x_\ell \in \Re^{d}$, define 
  the partial application:
\begin{align*}
  T(x_1, \cdots, x_\ell) 
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[i_1, i_2, \ldots, i_\ell] x_1[i_1] \cdots x_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell].
\end{align*}
%where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

@@

  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters:
\begin{align*}
  \mO_{ij} &= \sum_h {\pi \oft 1}[h] {\mOpp{i}{1}}[\cdot,h] \otimes {\mOpp{j}{1}}[\cdot,h] \quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOpp{1}{1}}[\cdot,h] \otimes {\mOpp{2}{1}}[\cdot,h] \otimes  {\mOpp{3}{1}}[\cdot,h].
\end{align*}

>>>>

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
Let $\Pa(a) \subset X \union H$ be the parents of a variable $a$ in a directed
  model, and 
let $\sN(a) \subset X \union H$ be the neighbors of $a$ in an
  undirected model.
We will use $x_1, x_2, \cdots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For an integer $n$, let $[n] = \{ 1, \dots, n \}$.
For ease of exposition, we assume that $x_v \in [d]$ for
all $v$, and that $h_i \in [k]$ for all $i$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef [d]^L$ and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef [k]^M$.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\bi
  = (i_1, \cdots, i_\ell)$, define the projection $T[\bi] = T[i_1 \ldots
  i_\ell]$.
Similarly, for a set $\sC$ of size $m$ and vector $\bx \in \Re^d$, we
  will use the notation $\bx_\sC$ to get a vector in $\Re^m$ which selects
  the appropriate elements of $\bx$.

We use $\otimes$ to denote the tensor product: if $u \in \Re^d,
  v \in \Re^k$, then $u \otimes v \in \Re^{d \times k}$.
For an $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $\bx_1, \cdots, \bx_\ell \in \Re^{d}$, define 
  the application:
\begin{align*}
  T(\bx_1, \cdots, \bx_\ell) 
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
&= \sum_{\bi = \ones}^{\vec d}
            T[i_1, i_2, \ldots, i_\ell] \bx_1[i_1] \cdots \bx_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell],
\end{align*}
a tensor in $\Re^{k \times \ldots \times k}$.
%where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

@@

  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters:
\begin{align*}
  \mO_{ij} &= \sum_h {\pi \oft 1}[h] {\mOppt{i}{1}}[h] \otimes {\mOppt{j}{1}}[h] \quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{1}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{3}{1}}[h].
\end{align*}

### 2014-05-10 doc/icml2014/directed.tex
Various minor corrections to camera-ready

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In general directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  do not correspond directly to the underlying parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to factor graphs;
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime; and
(iii) we present the algorithm solely in terms of linear algebra operations; in
  \sectionref{piecewise}, we show how we can get a more statistically
  efficient estimator using composite likelihoods.

@@

This property also extends to any connected sub-graph $\sG' \in \sG$.
\end{property}

For the rest of this section, we will only consider sub-graphs which are
  cliques. 
In \sectionref{undirected}, we will show how the generalization
  to sub-graphs allows us to efficiently estimate the pseudo-likelihood.

@@

% DONE: transition the content more gradually (checkpoint: what have we done so far, what next)
% DONE: refer to properties by name in addition to by number
We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}). But in a general graph, which cliques have this property?
In the example above, we were able to guarantee \propertyref{exclusive-views}
  by identifying a hidden variable, $h_1$ along with three conditionally
  independent observed variables, $x^a_1, x^b_1, x^a_2$.
  %followed by using \TensorFactorize to learn the conditional moments.
We define such a hidden variable to be a bottleneck:
\begin{definition}[Bottleneck]
  A hidden variable $h$ is said to be a \textbf{bottleneck} if there
  exists at least three observed variables, or views, $x_1, x_2, x_3$
  that are conditionally independent given $h$. 
  Let $\sV_{h}$ denote a set of views for $h$ (not necessarily unique).
\end{definition}

@@

% DONE: move here because not central to story
\paragraph{Remark.} We note that \propertyref{bottleneck} can be relaxed if some cliques
  share parameters.
In the directed grid model, we can identify the conditional moments $O$ from just
  $x^a_1, x^b_1$ and $x^a_2$.
  Therefore, we do not require that $h_2, h_3$ or $h_4$
  be bottlenecks, and can omit the observations $x^b_2, x^b_3$ and $x^b_4$.
%  \todo{make this a footnote if running out of space}
% PL: don't understand this comment
%In this case, we only require that the hidden variables in distinct
  %cliques share parameters.

@@

In general, $\sigma_{\min}(\mOpp{v}{i})$ can depend exponentially on the
  path length between the hidden variable $h_i$ and $x_v$.
Intuitively, this is expected because the influence from $h_i$ to $x_v$
  gets diluted for every latent variable it must factor through.

>>>>

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In general directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  do not correspond directly to the underlying parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to factor graphs;
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime; and
(iii) we present the algorithm solely in terms of linear algebraic operations; in
  \sectionref{piecewise}, we show how we can get a more statistically
  efficient estimator using composite likelihoods.

@@

This property also extends to any connected sub-graph $\sG' \in \sG$.
\end{property}
For the rest of this section, we will only consider sub-graphs which are
  cliques. 
In \sectionref{undirected}, we will show how the generalization
  to sub-graphs allows us to efficiently estimate the pseudo-likelihood.

@@

% DONE: transition the content more gradually (checkpoint: what have we done so far, what next)
% DONE: refer to properties by name in addition to by number
We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}). But in a general graph, which cliques have this property?
In the example above, we were able to guarantee \propertyref{exclusive-views}
  by identifying a hidden variable, $h_1$ along with three conditionally
  independent observed variables, $x^a_1, x^b_1, x^a_2$ using the tensor
  factorization method.
  %followed by using \TensorFactorize to learn the conditional moments.
We define such a hidden variable to be a bottleneck:
\begin{definition}[Bottleneck]
  A hidden variable $h$ is said to be a \textbf{bottleneck} if there
  exists at least three observed variables, or views, $x_1, x_2, x_3$
  that are conditionally independent given $h$. 
  Let $\sV_{h}$ denote a set of views for $h$ (not necessarily unique).
\end{definition}

@@

% DONE: move here because not central to story
\paragraph{Remarks.} We note that \propertyref{bottleneck} can be relaxed if some cliques
  share parameters.
In the directed grid model, we can identify the conditional moments $O$ from just
  $x^a_1, x^b_1$ and $x^a_2$.
  Therefore, we do not require that $h_2, h_3$ or $h_4$
  be bottlenecks, and can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

The method also extends directly to case in which the observed variables
  are real valued, while the hidden variables remain discrete. 
In this scenario, the tensor factorization method of
  \citet{anandkumar13tensor} recovers the expected conditional means,
  $\mu_{vi} \eqdef \E[x_v | h_i]$ for an observed variable $x_v$ and
  hidden variable $h_i$.
\algorithmref{learnclique} and \algorithmref{directed} still apply and allow
  us to recover clique marginals $Z_\sC$.
%  \todo{make this a footnote if running out of space}
% PL: don't understand this comment
%In this case, we only require that the hidden variables in distinct
  %cliques share parameters.

@@

In general, $\sigma_{\min}(\mOpp{v}{i})$ can depend exponentially on the
  path length between the hidden variable $h_i$ and $x_v$.
This is expected because the influence from $h_i$ to $x_v$
  gets diluted for every latent variable it must factor through.

### 2014-05-10 doc/icml2014/undirected.tex
Various minor corrections to camera-ready

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, the \emph{supervised} likelihood would be concave:
\begin{align}
\label{eqn:logLinearSupervised}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim \sD_\text{sup}}[\log p_\theta(\vx,\vh)] \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\end{align}
While we don't have supervised data,
note that the method of moments gives us hidden marginals $Z_\sC = \BP(\vh_\sC)$.
Furthermore, given $\mOpp{v}{i}$ and $Z_\sC$ of any $\sC \ni v$, we can recover
all marginals $\BP(\vx_\sC, \vh_\sC) = Z_\sC \cdot \mOpp[\vx_\sV]$\todo{clean up}.
Finally, the key step is that we can compute all the expected feature vectors from the marginals:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} \BP(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Therefore, we can optimize the supervised likelihood objective without actually
having supervised data.
Of course, in the finite data regime, the method of moments yields estimates
$\hat \mu^\text{mom}_\sC$ which approximate the true $\mu_\sC$.
In supervised learning, we have different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over the data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

@@

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{directed}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any graphical model satisfying the exclusive views property (\propertyref{exclusive-views}). 
Consequently, the result is that the output of \algorithmref{undirected},
$\hat\theta$, is a consistent estimator.\footnote{Of course, the log-linear model must be identifiable
(it is violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical).
In general, we can identify the parameters up to kernel of $\nabla^2 A(\theta)$.}

@@

However, pseudolikelihood also replaces the log-partition function $A(\theta)$
with a sum over the variables in the neighbourhood: $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)} \exp(\theta^\top\phi_{\sN(a)}(a'))$,
which cannot be computed from just the hidden clique marginals $Z_\sC$,
since the neighbors $\sN(a)$ in general do not form a clique.
We adapt our earlier algorithm ($\LearnClique$) to estimate the
  marginals over all neighborhoods $Z_{\sN(a)}$ by noting that
  \lemmaref{bottleneck-views} guarantees exclusive views for any connected
  sub-graph of $\sG$, not just cliques.

>>>>

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, the \emph{supervised} likelihood would be concave:
\begin{align}
\label{eqn:logLinearSupervised}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim \sD_\text{sup}}[\log p_\theta(\vx,\vh)] \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\end{align}
While we don't have supervised data,
note that the method of moments gives us hidden marginals $Z_\sC = \BP(\vh_\sC)$.
Furthermore, given $\mOpp{v}{i}$ and $Z_\sC$ of any $\sC \ni v$, we can recover
all marginals $\BP(\vx_\sC, \vh_\sC) = Z_\sC \cdot \mOppAll[\vx_\sV]$.%\todo{clean up}.
Finally, the key step is that we can compute all the expected feature vectors from the marginals:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} \BP(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Therefore, we can optimize the supervised likelihood objective without actually
having supervised data.
Of course, in the finite data regime, the method of moments yields estimates
$\hat \mu^\text{mom}_\sC$ which approximate the true $\mu_\sC$.
In supervised learning, we have different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over the data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

@@

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{directed}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any graphical model satisfying the exclusive views property (\propertyref{exclusive-views}). 
Consequently, the result is that the output of \algorithmref{undirected},
$\hat\theta$, is a consistent estimator.\footnote{Of course, the log-linear model must be identifiable
(it is violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical).
In general, we can identify the parameters up to the kernel of $\nabla^2 A(\theta)$.}

@@

The pseudolikelihood also replaces the log-partition function $A(\theta)$
with a sum over the variables in the neighbourhood: $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)} \exp(\theta^\top\phi_{\sN(a)}(a'))$,
which cannot be computed from just the hidden clique marginals $Z_\sC$,
since the neighbors $\sN(a)$ in general do not form a clique.
We adapt our earlier algorithm ($\LearnClique$) to estimate the
  marginals over all neighborhoods $Z_{\sN(a)}$ by noting that
  \lemmaref{bottleneck-views} guarantees exclusive views for any connected
  sub-graph of $\sG$, not just cliques.
